---
title: "Codebook - Aggregation of Human Activity Recognition Using Smartphones"
author: "Mike Quashne"
date: "1/2/2020"
output:
    md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Columns included in this dataset
The following columns are included:

* Subject --- Integer Factor, 1 to 30. The number assigned to a specific smartphone from which observations were taken.


* Activity --- String Factor. The motion activity being performed at the time of the observation. Motion activities are:
  + WALKING
  + WALKING_UPSTAIRS
  + WALKING_DOWNSTAIRS
  + SITTING
  + STANDING
  + LAYING
  
  
* Numbered Summary Statistics (remaining 66 columns) --- Double, -1 to 1. Each column contains an average of the named variable for a given Subject and Activity. These variables represent a summary statistic (mean or standard deviation) related to standardized and normalized measurements from a smartphone accelerometer at the time of observation. The variables included as columns in this dataset are:
  + 1 tBodyAcc-mean()-X
  + 2 tBodyAcc-mean()-Y
  + 3 tBodyAcc-mean()-Z
  + 41 tGravityAcc-mean()-X
  + 42 tGravityAcc-mean()-Y
  + 43 tGravityAcc-mean()-Z
  + 81 tBodyAccJerk-mean()-X
  + 82 tBodyAccJerk-mean()-Y
  + 83 tBodyAccJerk-mean()-Z
  + 121 tBodyGyro-mean()-X
  + 122 tBodyGyro-mean()-Y
  + 123 tBodyGyro-mean()-Z
  + 161 tBodyGyroJerk-mean()-X
  + 162 tBodyGyroJerk-mean()-Y
  + 163 tBodyGyroJerk-mean()-Z
  + 201 tBodyAccMag-mean()
  + 214 tGravityAccMag-mean()
  + 227 tBodyAccJerkMag-mean()
  + 240 tBodyGyroMag-mean()
  + 253 tBodyGyroJerkMag-mean()
  + 266 fBodyAcc-mean()-X
  + 267 fBodyAcc-mean()-Y
  + 268 fBodyAcc-mean()-Z
  + 345 fBodyAccJerk-mean()-X
  + 346 fBodyAccJerk-mean()-Y
  + 347 fBodyAccJerk-mean()-Z
  + 424 fBodyGyro-mean()-X
  + 425 fBodyGyro-mean()-Y
  + 426 fBodyGyro-mean()-Z
  + 503 fBodyAccMag-mean()
  + 516 fBodyBodyAccJerkMag-mean()
  + 529 fBodyBodyGyroMag-mean()
  + 542 fBodyBodyGyroJerkMag-mean()
  + 4 tBodyAcc-std()-X
  + 5 tBodyAcc-std()-Y
  + 6 tBodyAcc-std()-Z
  + 44 tGravityAcc-std()-X
  + 45 tGravityAcc-std()-Y
  + 46 tGravityAcc-std()-Z
  + 84 tBodyAccJerk-std()-X
  + 85 tBodyAccJerk-std()-Y
  + 86 tBodyAccJerk-std()-Z
  + 124 tBodyGyro-std()-X
  + 125 tBodyGyro-std()-Y
  + 126 tBodyGyro-std()-Z
  + 164 tBodyGyroJerk-std()-X
  + 165 tBodyGyroJerk-std()-Y
  + 166 tBodyGyroJerk-std()-Z
  + 202 tBodyAccMag-std()
  + 215 tGravityAccMag-std()
  + 228 tBodyAccJerkMag-std()
  + 241 tBodyGyroMag-std()
  + 254 tBodyGyroJerkMag-std()
  + 269 fBodyAcc-std()-X
  + 270 fBodyAcc-std()-Y
  + 271 fBodyAcc-std()-Z
  + 348 fBodyAccJerk-std()-X
  + 349 fBodyAccJerk-std()-Y
  + 350 fBodyAccJerk-std()-Z
  + 427 fBodyGyro-std()-X
  + 428 fBodyGyro-std()-Y
  + 429 fBodyGyro-std()-Z
  + 504 fBodyAccMag-std()
  + 517 fBodyBodyAccJerkMag-std()
  + 530 fBodyBodyGyroMag-std()
  + 543 fBodyBodyGyroJerkMag-std()
  
# Data Description from source files
The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

* tBodyAcc-XYZ
* tGravityAcc-XYZ
* tBodyAccJerk-XYZ
* tBodyGyro-XYZ
* tBodyGyroJerk-XYZ
* tBodyAccMag
* tGravityAccMag
* tBodyAccJerkMag
* tBodyGyroMag
* tBodyGyroJerkMag
* fBodyAcc-XYZ
* fBodyAccJerk-XYZ
* fBodyGyro-XYZ
* fBodyAccMag
* fBodyAccJerkMag
* fBodyGyroMag
* fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

* mean(): Mean value
* std(): Standard deviation


# Data Manipulation

## Downloading Data
Data for this set was downloaded as a zip file from <https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip>

This zip file contains observations from the accelerometers of 30 Samsung Galaxy S smartphones during six different motion activities.

The following files were used to compose this dataset:

File|Description
--|--
activitylabels.txt|A lookup table providing text for each of the 6 numbered activities in the datasets.
features.txt|A list of the names of 561 variables gathered from the accelerometer data. These measurements correspond to the columns in the X datasets.
subject_test.txt|A column of integers representing which of the 30 test subjects each observation in the X_test dataset came from.
subject_train.txt|A column of integers representing which of the 30 test subjects each observation in the X_train dataset came from.
Y_test.txt|A column of integers representing which of the six motion action activities corresponds to each observation in the X_test dataset
Y_train.txt|A column of integers representing which of the six motion action activities corresponds to each observation in the X_train dataset
X_text.txt|A table of measurements containing 561 variables for 2,947 discrete test observations from subjects during motion activities. The variables correspond to the names in features.txt
X_train.txt|A table of measurements containing 561 variables for 7,352 discrete training observations from subjects during motion activities. The variables correspond to the names in features.txt

The files were downloaded and stored in a folder in the working directory.



## Requirement #1 - Merge the training and test sets
The following packages were loaded for use in the analysis:
```{r libraries_loaded,warning=FALSE,message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
```

Then, each text file was read into an r tibble variable using read_table from the dplyr package. No files contained column names, so the col_names option was set to false:
```{r files_read,warning=FALSE,message=FALSE}
subject_test<-read_table("./UCI HAR Dataset/test/subject_test.txt", col_names=FALSE)
X_test<-read_table("./UCI HAR Dataset/test/X_test.txt", col_names=FALSE)
Y_test<-read_table("./UCI HAR Dataset/test/Y_test.txt", col_names=FALSE)
subject_train<-read_table("./UCI HAR Dataset/train/subject_train.txt", col_names=FALSE)
X_train<-read_table("./UCI HAR Dataset/train/X_train.txt", col_names=FALSE)
Y_train<-read_table("./UCI HAR Dataset/train/Y_train.txt", col_names=FALSE)
activity_labels<-read_table("./UCI HAR Dataset/activity_labels.txt", col_names=FALSE)
features<-read_table("./UCI HAR Dataset/features.txt", col_names=FALSE)
```

After reading in the data, the tibbles were formatted for a clean merge. The merge command rearranges the observations included in each dataset, so much of the pre-processing occurs before the merge.

First, activities in Y_test and Y_train were bound to the corresponding observations in X_test and X_train, respectively, and the column was named "Activity" in both tibbles:
```{r activities_bound}
colnames(Y_test)<-"Activity"
colnames(Y_train)<-"Activity"
X_test<-tbl_df(cbind(Y_test,X_test))
X_train<-tbl_df(cbind(Y_train,X_train))
```

Then, the corresponding subject from subject_test and subject_trained were bound to each observation in X_test and X_train, and the column named "Subject":
```{r subject_bound}
colnames(subject_test)<-"Subject"
colnames(subject_train)<-"Subject"
X_test<-tbl_df(cbind(subject_test,X_test))
X_train<-tbl_df(cbind(subject_train,X_train))
```

Prior to the merge, a column (Observation_Type) was added to each dataset using dplyr's mutate function with a string corresponding to which of the observation files the observation came from. "Test" was used for observations in X_test and "Train" was used for observations in X_train. This column was included to allow for later decomposition of the dataset if necessary:
```{r observation_type}
X_test<-mutate(X_test,Observation_Type = "Test")
dim(X_test)
X_train<-mutate(X_train,Observation_Type = "Train")
dim(X_train)
```


Then, with each tibble formatted the same way, the X_test and X_train tibbles were merged into a single tibble called X_set using option all=TRUE to preserve all observations from both tables:

```{r merge}
X_set<-tbl_df(merge(X_test,X_train,all=TRUE))
dim(X_set)
```


**This fulfills assignment criteria #1**



## Requirement #4 - Descriptive variable names
Following the merge, a few cleanup operations were performed. First, column names were added to the X_set using already defined column names and a character vector made up of the column from the features tibble:

```{r cleanup}
colnames(X_set)<-c("Subject","Activity",unlist(features[,1]),"Observation_Type")
```
**This fulfills assignment criteria #4**



## Requirement #3 - Descriptive activity names

Then, the "Subject", "Activity", and "Observation_Type" columns were converted into factors from double, double, and character, respectively:
```{r column_classes}
X_set<-mutate(X_set,Subject=as.factor(Subject),Activity=as.factor(Activity), Observation_Type=as.factor(Observation_Type))
```

Finally, the factor numbers in the "Activity" column were replaced with the corresponding text from the activity_lables tibble:

```{r activity_labels}
levels(X_set$Activity)<-unlist(activity_labels[,2])
```
**This fulfills assignment criteria #3**

The result is a fully merged dataset with descriptive variable names and activity names (requirements 1, 3, and 4).



## Requirement #2 - Select only certain columns

To create the final product, only the factor variable columns, as well as those containing the mean and standard deviation of individual observations are needed. The mean and standard deviation columns all end in either "mean()" or "std()". A select function was performed to gather the factor variables (excluding Observation_Type), and any columns containing "mean()" or "std()", bringing the total number of columns from 564 to 68: 

```{r downselect}
X_agg<-select(X_set,Subject,Activity,contains("mean()"),contains("std()"))
dim(X_agg)
```


**This fulfills assignment criteria #3**


## Requirement #5 - Create a second dataset with the average of each variable for each activity and each subject

To create the final dataset, the aggregated set from the previous step was grouped by subject and activity, then the summarize_all function in dplyr was used to find the mean of all non-grouped variables:

```{r summary}
Summary_Set<-group_by(X_agg,Subject,Activity) %>% summarize_all(funs(mean))
write.table(Summary_Set,"Dataset.txt")
Summary_Set
```


**This fullfils assignment criteria #5, and finishes the assignment**


